Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-11-13 output/breed_paper.Rnw
Final edits.

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using 33 years of experimental
yield data from the California Cooperative Rice Research Foundation
Rice Experiment Station. Based on side-by-side comparisons of old and
new rice varieties which do not consider yield erosion, there was an
apparent increase in yield. However, the yields of older varieties
eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1}
(90\% credible interval -4.4 to -53.3) after initial selection. Once
this effect was considered, the yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). These results highlight first, the
importance of continuous crop improvement and deployment of new
cultivars simply to maintain existing yields, and second, to increase
the genetic yield potential, higher yield targets are
needed. Importantly, when breeding near the yield potential, despite
the limited yield gains, significant advances in improving quality and
reducing crop duration have been made.

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area~\citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield~\citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley, which is characterized by
having a Mediterranean climate with long days, a dry growing season,
and relatively free of pests and diseases. These conditions lead to
some of the highest yields in the world~\citep{FAO2016}. Production in
CA is predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
year\tsups{-1} between 1999 and 2014.

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluate current and
promising cultivars at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials per
year, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top 15cm~\citep{ssurgo}.

@@

To test the hypothesis, the influence of breeding and yield erosion
were quantified on yield over time using a Bayesian hierarchical
model. Yield was modeled as a response to the number of years since
the cultivar was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between
years and between cultivars, year and cultivar were included in the
model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the
$Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit to estimate
if there were differences between cultivars in the rate of yield
erosion via a random slope by cultivar (Model 3).

Data were processed and models fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval~\citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods, specifically a frequentist mixed-effects
model fit by maximum likelihood (see supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

# text(x = plot_data$yr, y = plot_data$mean,
#      labels = 1:nrow(plot_data), #gsub("M", "M-", row.names(plot_data)),
#      pos = 4, cex = 0.75)
# mm = lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,])
# abline(mm, lty = 2)
# yrs = 1980:2020
# int = predict(mm, newdata = data.frame(release_yr = yrs), se.fit = TRUE)

# polygon(x = c(yrs, rev(yrs)),
#         y = c(int$fit + int$se.fit,rev(int$fit - int$se.fit)),
#         border = NA,
#         col = rgb(0,0,0,0.1))

@@

There was general agreement between the estimates from the Bayesian
and classical approaches for all models, with the estimates being
similar between approaches for similar models. The Bayesian estimates
are presented here due to the inherent flexibility of Bayesian
inference, though the classical estimates can be found in the
supplemental material. Comparing the cultivars' yield during each's
initial time (average 3 years) in the experimental trials just before
being released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
provides evidence of genetic yield improvements during the 33-year
study period. Individual exceptions to this were two cultivars (M-204
and M-205) that had higher yields than the others.

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3), hence the model excluding these
effects (Model 2) is presented here. The yields of each cultivar over
time were evaluated relative to when the cultivar was released using
the full model which accounts for both yield erosion and yield
improvements over time. The average decrease in yield over time (i.e.,
yield erosion) once a cultivar was selected was estimated at
\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
\ref{tbl:coef}). The oldest cultivar in the trials
(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
is estimated to have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline. M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed
solely to the erosion of M-202's yield over time (Figs
\ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further supporting this
estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}; Table \ref{tbl:ids}).

@@

When yield erosion is not accounted for in the analysis (i.e.,
omitted from the model), there is estimated to be an increase per year
of approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
full model which accounts for both the year of release and yield
erosion indicates that the estimated change in grain yield per year
in the CCRRF's breeding trials at the RES was \Sexpr{release_yr_coef}
kg ha\tsups{-1} year\tsups{-1}, with both small positive and negative
trends within the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}).

@@

# par(mfrow = c(1,2),
#     mar = c(4,0,2,0.75),
#     oma = c(4,4,0.5,0.5))

@@

In California, the cultivars with broad adaptation and adoption over
the 33 year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in these cultivars at any given time during the study
period~\citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
and M-402 which is a premium quality cultivar. Other cultivars such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) that does not account for yield erosion is
lower than the 37 to 60 kg ha\tsups{-1} year\tsups{-1} estimated
by~\citet{Omar2014} in side-by-side comparisons. However, when
accounting for yield erosion and looking acrossall cultivars, there is
no conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this cultivar is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or cooler
 temperatures reducing yields. Again, this is
unlikely to be the case in this situation as ~\citet{espe2017} found
no significant changes in growing season temperatures (either high,
low or average temps) the Sacramento Valley during the study period.
Thirdly, biotic and abiotic forces have previously been speculated as
the cause for yield erosion of the cultivar IR8 at
IRRI~\citep{Peng1998, Peng2000, Peng2010}. They argued that since the
cultivar's genetics remain stable from the time of selection but the
conditions the cultivar experiences over time are continuously
changing leading ot yield erosion~\citep{Peng1998, Peng2000,
Peng2010}. However, it is not clear what is responsible for the yield
erosion of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in
this analysis. It is unlikely biotic factors as pest and disease
pressure are low in this Mediterranean climate. Furthermore, these
fields are fallowed every other year and the residues are burned,
further lowering the potential for pest and disease pressure. While
efforts should be made to identify the cause of erosion in this
environment, this observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield
~\citep{espe2017} potential, especially in systems such as CA where yield is approaching
the physiological limit \citep{espe16a}. Without maintenance breeding
(i.e., continuously selecting cultivars well adapted to current
conditions), CA rice production systems are predicted to have lost an
estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
yield over the study period.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields~\citep{espe2017}, and the RES experiences warmer
temperatures than many other rice growing areas in CA
\citep{espe2017}. Hence, cultivars which perform well at the RES might
not perform well across of CA. Since these stressful temperature
events are inconsistent across years, they influence yield stability (how consistent
yield is year to year). Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality~\citep{mckenzie1993} and
milling yield~\citep{mckenzie1994}, have been made over the study
period and are important. Maintaining high quality is imperative for
the breeding program and it is not uncommon in the breeding program
for higher yielding cultivars to be eliminated due to concerns about
quality. However, the goal of increasing yield may require the yield
performance of cultivars to be assessed earlier in the selection
process to avoid discarding these higher yielding entries without full
consideration. Another change in CA rice cultivars is that crop
duration has been reduced. For example, during this study period,
M-202 and M-206 have been the predominate cultivars grown in CA, and
while their yield potentials were similar at the time of release
(Fig. 5), M-206 matures about 6 days earlier than M-202 (Table
\ref{tbl:ids}). Shorter duration is very important in CA due to an
already short growing season and water limitations (shorter duration
cultivars would presumably require less irrigation water).

>>>>

Meeting the challenge of feeding a growing population with limited
resources will require increasing the yield potential of staple crops,
such as rice. Yet many high-yielding, intensive production systems
have experienced slow rates of yield improvement in recent years
despite a demonstrated increase in the yield potential of new crop
cultivars. We analyzed experimental data from one such cropping
system, i.e., California (CA) rice, in order to quantify improvements
made in the genetic yield potential obtained through plant
breeding. California rice systems are among the highest in the world
and close to maximum yield potential. Specifically, the hypothesis was
tested that if rice cultivar yields decline over time (termed yield
erosion) then apparent yield increases in side-by-side yield
comparison tests will not reflect increases in yield potential. This
hypothesis was tested using 33 years of experimental yield data from
the California Cooperative Rice Research Foundation Rice Experiment
Station. Based on side-by-side comparisons of old and new rice
cultivars which do not consider yield erosion, there was an apparent
increase in yield. However, the yields of older cultivars eroded at an
estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible
interval -4.4 to -53.3) after initial selection. Once this effect was
considered, the yield advantage of newer cultivars over old was
uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
-36.1 to 31.5). These results highlight 1) the importance of
continuous crop improvement and deployment of new cultivars simply to
maintain existing yields, and 2) to increase the genetic yield
potential, higher yield targets are needed. Importantly, when breeding
near the yield potential, despite the limited yield gains, significant
advances in improving quality and reducing crop duration have been
made.

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area~\citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield~\citep{Grassini2013}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley, which is characterized by
having a Mediterranean climate with long days, a dry growing season
relatively free of pests and diseases. These conditions lead to some
of the highest yields in the world~\citep{FAO2016}. Production in CA
is predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
year\tsups{-1} between 1999 and 2014.

Certainly one question in looking at this situation is, ``do new
cultivars have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluate current and
promising cultivars at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials
between years, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top
15cm~\citep{ssurgo}.

@@

To test the hypothesis, the influence of breeding and yield erosion
were quantified on yield over time using three nested Bayesian hierarchical
models. First, yield was modeled as a response to the number of years since
the cultivar was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between
years and between cultivars, year and cultivar were included in the
model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. This
full model was modified to further explore yield dynamics over
time. For comparison to estimates where yield erosion is not taken
into account, the model was used as described above except omitting
the $Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit
to estimate if there were differences between cultivars in the rate of
yield erosion by adding a random slope by cultivar (Model 3).

Data were processed and models fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to sample to sample variance
and hence is preferable to the 95\% interval~\citep{rstanarm}. To
verify the estimates, the analysis was also conducted using classical
methods, specifically frequentist mixed-effects models fit by maximum
likelihood (see supplemental material for details and model
results). Complete data and code used for this analysis is available
at \url{github.com/mespe/rice_advances.git}.

@@

@@

There was general agreement between the estimates from the Bayesian
and classical statistical approaches for all models, with the
estimates being similar between similar models. The Bayesian estimates
are presented here due to the inherent flexibility of Bayesian
inference, though the classical estimates can be found in the
supplemental material. Comparing the cultivars' yield during each's
initial time (average 3 years) in the experimental trials just before
being released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Comparing the yields from only the first few
years for each cultivar over the study period does not provide
evidence of genetic yield improvements during the 33-year study
period. Individual exceptions to this were two cultivars (M-204 and
M-205) that had higher yields than the others.

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3), hence the model excluding varying
rates of erosion by cultivar (Model 2) is presented here. The yields
of each cultivar over time were evaluated relative to when the
cultivar was released using the full model which accounts for both
yield erosion and yield improvements over time. The average decrease
in yield over time (i.e., yield erosion) once a cultivar was selected
was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
Table \ref{tbl:coef}). The oldest cultivar in the trials
(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
is estimated to have lost an average of
\Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since its
release. Likewise, an estimated 560 kg ha\tsups{-1} yield
advantage of M-206 over M-202 (the current and former predominant
cultivars, respectively), can be attributed solely to the erosion
of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:ids}).

@@

\subsection{Yield gains}

When yield erosion is not accounted for in the analysis (Model 2),
there is estimated to be an increase per year of approximately
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
\Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
accounts for both the year of release and yield erosion indicates that
the estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
year\tsups{-1}, with both small positive and negative trends within
the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
estimated to have superior predictive accuracy compared to the model
which does not account for yield erosion (Model 2)

@@

@@

In CA, the cultivars with broad adaptation and adoption over the 33
year period of this study are M-202, M-206, and M-105. Cold tolerance
is a major problem for much of the California rice producing region
~\citep{espe2017}, although it is a bigger problem in the southern
portion of the Sacramento Valley due to the influence of the ocean
winds through the Delta ~\citep{espe16a}. These three cultivars have a
higher degree of cold tolerance, high yield potential, and relatively
short duration and are thus broadly grown with over 70\% of the area
in these cultivars at any given time during the study
period~\citep{ucce}. There are several cultivars that were found to
have better than average yields, even after correcting for yield
erosion. However, these cultivars (M-201, M-204, M-205, and possibly
M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
not all medium grain cultivars that were released were purported to
have higher yields than previous cultivars. Some cultivars had special
traits, but upon release had similar yields to more broadly adapted
cultivars. Examples of this include M-208, which had resistance to
blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
premium quality cultivar. Other cultivars such as M-205 and M-209,
while high yielding, do not have the cold tolerance of the other
cultivars and are thus only recommended for the northern portion of
the Sacramento Valley.

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. However, when accounting for yield erosion and looking
across all cultivars, there is no conclusive evidence of increased
genetic yield potential in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
Grain yield over this period has stayed relatively constant at an average
\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}).

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Second, changing climate may influence yields
with either warmer~\citep{van2013} or cooler~\citep{espe2017}
temperatures reducing yields. Again, this is unlikely to be the case
in this situation as ~\citet{espe2017} found no significant changes in
growing season temperatures (either high, low, or average temps) the
Sacramento Valley during the study period. Thirdly, small changes in
biotic and abiotic forces have previously been speculated as the cause
for yield erosion of the cultivar IR8 at IRRI~\citep{Peng1998,
Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar experiences over
time are continuously changing, accumulated changes too small to
reliably measure could plausibly lead to yield
erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
if this mechanism is responsible for the yield erosion of
\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure.

It is striking that the estimated yield erosion is similar to that
reported by \citet{Peng2010} over a similar time span (15\% decline
over roughly 30 years). This suggests the mechanisms responsible could
potentially be endemic to all rice production systems. While efforts
should be made to identify the cause of erosion, this study highlights
both the need for maintenance of plant breeding and the difficulty of
increasing genetic yield ~\citep{espe2017} potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}.

@@

Our investigation here is narrowly focused on the rough-rice/field
grain yield. Other advancements due to plant improvement, including
increased quality~\citep{mckenzie1993} and milling
yield~\citep{mckenzie1994}, have been made over the study period and
are important. Additionally, crop duration has been reduced in CA rice
cultivars. For example, while the predominate cultivars M-202 and
M-206 had similar yield potentials at the time of release (Fig. 5),
M-206 matures about 6 days earlier than M-202 (Table
\ref{tbl:ids}). Shorter duration is very important in CA due to an
already short growing season and water limitations (shorter duration
cultivars would presumably require less irrigation water).
